---
title: 'Assginment_02'
author: 'Yingtong Zhang'
date: "Jan 31, 2019"
output: html_document
---

## Questions:

1. Generate plots of the logistic growth model at r = 1.95, 2.05, 2.5, and 2.8 Describe the trajectory observed in each case.

```{r}
r1 = 1.95     ## intrinsic growth rate
r2 = 2.05
r3 = 2.5
r4 = 2.9
K = 10        ## carrying capacity      
n0 = .1       ## initial population size
NT = 30       ## number of time steps to simulate
time = 1:NT
```

Iterative simulation
```{r}
n1 = n2 = n3 = n4 = rep(n0,NT)    ## vector to store results
for(t in 2:NT){
  n1[t] = n1[t-1] + r1*n1[t-1]*(1-n1[t-1]/K)
  n2[t] = n2[t-1] + r2*n2[t-1]*(1-n2[t-1]/K)
  n3[t] = n3[t-1] + r3*n3[t-1]*(1-n3[t-1]/K)
  n4[t] = n4[t-1] + r4*n4[t-1]*(1-n4[t-1]/K)
}
```

Plot results
```{r}
plot(time,n1,ylim=c(0,12),lwd=3,type='l',
     bty='l',cex.lab=1.5,
     xlab="Time",ylab="Population Size")

lines(time, n2, lwd=2, col=2)
lines(time, n3, lwd=1, col=3)
lines(time, n4, lwd=1, col=8)
```
```
The variation in r dipicts how the system shifts from stable to periodic oscillations to chaotic as r increases, which is a more random trajectory when r increased from 1.95 to 2.9. It illustrated that discrete time logistic
model could produce chaotic dynamics.
```

2. Choose another probability distribution and generate graphs of the probability density function, the cumulative distribution function, the quantile function, and a histogram of samples from that distribution.
```{r}
# take poisson distribution for example
# PDF
x = seq(0,20)
plot(x,dpois(x,1), type='l')     
lines(x,dpois(x,4), col=2)           ## Poisson distribution has only one parameter -- lamda
lines(x,dpois(x,10), col=3)
legend(16, 0.35, legend=c("位 = 1", "位 = 4", "位 = 10"),           # add the legend
       col=c("black", "red", "green"), lty=1:2, cex=0.8)


# CDF
plot(x,ppois(x,1),type='l',ylim=c(0, 1))
lines(x,ppois(x,4),col=2)
lines(x,ppois(x,10),col=3)
legend(16, 0.25, legend=c("位 = 1", "位 = 4", "位 = 10"),        
       col=c("black", "red", "green"), lty=1:2, cex=0.8)


# quantile
p = seq(0,1,by=0.1)
plot(p,qpois(p,1),type='l',ylim=range(x))
# range returns the min/max as a 2-element vector
lines(p,qpois(p,4),col=2)
lines(p,qpois(p,10),col=3)
legend(0.8, 18, legend=c("位 = 1", "位 = 4", "位 = 10"),           # add the legend
       col=c("black", "red", "green"), lty=1:2, cex=0.8)

```
For Poisson, the mean and the variance are both  .The 95-percent confidence interval is
```{r}
#a = 4 - 1.96 * sqrt(4/10)
#b = 4 + 1.96 * sqrt(4/10)
qpois(c(0.025,0.975),4)

plot(p,qpois(p,4),type='l',ylim=range(x))
abline(v=c(0.025,0.975),lty=2)  
abline(h=qpois(c(0.025,0.975),4),lty=2)

plot(x,dpois(x,1), type='l') 
abline(v=qpois(c(0.025,0.975),4),lty=2)

# random
n = c(10,100,1000,10000)    # sequence of sample sizes
for(i in 1:4){          # loop over these sample sizes
  hist(rpois(n[i], 4),main=n[i],probability=TRUE,breaks=40)  
                #here breaks defines number of bins in the histogram
  lines(x,dpois(x, 4),col=2)
}
```

3. Numerically transform a lognormal(meanlog=0,sdlog=0.5) through sin(x) using Monte Carlo simulation. Include histograms of the original and transformed distributions. Report the mean, median, and 95% CI for both distributions and indicate these values on the histograms.

```{r}
x = rlnorm(1000,0,0.5)
y = sin(x)

hist(x,main="Original distribution",breaks=40)
abline(v=quantile(x,c(0.025,0.5,0.975)),lty=c(2,1,2),lwd=3,col="orange")
abline(v=mean(x),col="red",lwd=3,lty=3)
mean(x)
quantile(x,c(0.025,0.5,0.975))

hist(y,main="Transformed distribution",breaks=40)
abline(v=quantile(y,c(0.025,0.5,0.975)),lty=c(2,1,2),lwd=3,col="orange")
abline(v=mean(y),col="red",lwd=3,lty=3)
mean(y)
quantile(y,c(0.025,0.5,0.975))
```
```
The mean of the original distribution is 1.12148; the median is 0.9941163 and the 95% CI is 0.3418580, 2.7226157. The Monte Carlo estimate is that the mean is 0.7448914, the median is 0.7960687 and the 95% CI is 0.2467068, 0.9993069.

```

4. Plot histograms of the samples of r and K used for the simulation.
```{r}
r.sd = 0.2     ## standard deviation on r
K.sd = 1.0     ## standard deviation on K
NE = 1000      ## Ensemble size

r = 1
n = matrix(n0,NE,NT)
rE = rnorm(NE,r,r.sd)  
KE = rnorm(NE,K,K.sd) 

hist(rE, probability=TRUE,breaks=40)  
hist(KE, probability=TRUE,breaks=40) 
```

5. Plot a sample of 10 different trajectories (through time) from your ensemble (on one graph).

```{r}
for(i in 1:NE){        # loop over samples
  for(t in 2:NT){      # for each sample, simulate throught time
    n[i,t] = n[i,t-1] + rE[i]*n[i,t-1]*(1-n[i,t-1]/KE[i])
  }
}

plot(time,n[100,],lwd=3,ylim=c(0,12),type='l',bty='l',cex.lab=1.5,
     xlab="Time",ylab="Population Size")

for (i in 2:10){
  lines(time, n[i*100,],lty=2,type='l',bty='l',cex.lab=1.5)
}

```

6. Plot a histogram of your population forecast at time = 15.
```{r}
hist(n[,15],lwd=2,main='Histogram at time = 15',probability=TRUE,breaks=40)

```

7. Plot the median trajectory through time. Use ecoforecastR::ciEnvelope to add a 95% CI (i.e. 2.5% to 97.5%) to the plot. This function need to take time-series for both the upper (yhi) and lower (ylo) intervals.
```{r}
ciEnvelope <- function(x,ylo,yhi,col="lightgrey",...){
      polygon(cbind(c(x, rev(x), x[1]), c(ylo, rev(yhi),
      ylo[1])), border = NA,col=col,...)
}



n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
ylo = n.stats[1,]
yhi = n.stats[3,]
yme = n.stats[2,]

plot(time, yme, ylim=c(0,12),lty=2,lwd=3,col="orange")

ciEnvelope(time, ylo=ylo, yhi=yhi,col="lightgrey")

points(time, yme ,ylim=c(0,12),lty=2,lwd=3,col="orange")

```

Extra Credit: Initial conditions
--------------------------------

The approach for simulating uncertainty in the initial conditions is very similar to the approach used for the parameter uncertainty.  As in Chapter 2, we'll assume that the initial condition is distributed as a lognormal to ensure that we never draw negative values. For this example we'll assume a standard deviation of 0.6 and an intrinsic growth rate of 0.3

```{r}
r = 0.3
n0.sd = 0.6
n0s = rlnorm(NE,log(n0),n0.sd)
n = matrix(n0s,NE,NT)
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}
```

### Problems

8. Plot the median & 95% interval.
```{r}

n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
ylo = n.stats[1,]
yhi = n.stats[3,]
yme = n.stats[2,]

plot(time, yme, ylim=c(0,12),lty=2,lwd=3,col="orange")

ciEnvelope(time, ylo=ylo, yhi=yhi,col="lightgrey")

points(time, yme ,ylim=c(0,12),lty=2,lwd=3,col="orange")
```

9. Repeat with r equal to 1.95, 2.05, and 2.8
```{r}
r = 1.95
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}

n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
ylo = n.stats[1,]
yhi = n.stats[3,]
yme = n.stats[2,]

plot(time, yme, ylim=c(0,12),lty=2,lwd=3,col="orange")
ciEnvelope(time, ylo=ylo, yhi=yhi,col="lightgrey")
points(time, yme ,ylim=c(0,12),lty=2,lwd=3,col="orange")


r = 2.05
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}

n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
ylo = n.stats[1,]
yhi = n.stats[3,]
yme = n.stats[2,]

plot(time, yme, ylim=c(0,12),lty=2,lwd=3,col="orange")
ciEnvelope(time, ylo=ylo, yhi=yhi,col="lightgrey")
points(time, yme ,ylim=c(0,12),lty=2,lwd=3,col="orange")

r = 2.8
for(i in 1:NE){
  for(t in 2:NT){
    n[i,t] = n[i,t-1] + r*n[i,t-1]*(1-n[i,t-1]/K)
  }
}

n.stats = apply(n,2,quantile,c(0.025,0.5,0.975))
ylo = n.stats[1,]
yhi = n.stats[3,]
yme = n.stats[2,]

plot(time, yme, ylim=c(0,12),lty=2,lwd=3,col="orange")
ciEnvelope(time, ylo=ylo, yhi=yhi,col="lightgrey")
points(time, yme ,ylim=c(0,12),lty=2,lwd=3,col="orange")



```



