---
title: "Assginment_09_Ex09_Yingtong"
author: "Yingtong Zhang"
date: "4/1/2019"
output: html_document
---
Kalman Filter
========================================================

### Assignment

The assignment is to run the KF under all four combinations of covariance in the process model versus process error and compare the results. In particular you'll want to pay attention to the missing data at the beginning of the timeseries for some states. You'll also want to comment on how spatial adjacency affects the confidence in the inferences (some states are more isolated than others) in the four different scenarios. Finally, you'll want to note that the alpha estimated from the data itself (0.000209), is close to zero and thus our real forecast would be much more like our no-flux run than our high flux run.


```{r, echo=FALSE}
#devtools::install_github("EcoForecast/ecoforecastR")
library(ecoforecastR)
```



In this assignment we'll assimilate all 4 combinations of with/without interactions in the process model versus with/without correlation in the process error to evaluate how each impacts the inferences made.

```{r}
## load the Google flu data & select states
gflu = read.csv("http://www.google.org/flutrends/about/data/flu/us/data.txt",skip=11)
time = as.Date(gflu$Date)
states = c("Massachusetts","Connecticut","Rhode.Island","New.Hampshire","Vermont","Maine")
nstates = length(states)
y = t(gflu[,states])

## plot time-series from states
plot(time,1:length(time),type='n',ylab="Flu Index",lwd=2,log='y',ylim=range(y,na.rm=TRUE))
for(i in 1:nstates){
  lines(time,y[i,],col=i,lwd=2)
}
legend("topleft",legend=states,lwd=2,col=1:nstates)
```

```{r}
## define adjacency between states slected
adj = matrix(c(0,1,1,1,1,0,    ### state-to-state spatial adjacency (self=0)
               1,0,1,0,0,0,
               1,1,0,0,0,0,
               1,0,0,0,1,1,
               1,0,0,1,0,0,
               0,0,0,1,0,0),nstates,nstates,byrow=TRUE)
```

Thus, if state j has more cases than state i, this will tend to increase infection in state i. For your reference, below is the JAGS model fit to the log-transformed flu data

```{r}
SpatialRandomWalk = "
model{

  #### Data Model
  for(t in 1:n){
    for(i in 1:nstate){
      y[i,t] ~ dnorm(x[i,t],tau_obs)
    }
  }

  #### Process Model
  for(t in 2:n){
    for(i in 1:nstate){
      mu[i,t] <- x[i,t-1] + alpha * sum(adj[i,1:nstate]*x[1:nstate,t-1])
    }
    x[1:nstate,t] ~ dmnorm(mu[1:nstate,t],Omega_proc)
  }
  
  #### Priors
  for(i in 1:nstate){
    x[i,1] ~ dnorm(x_ic,tau_ic)
  }
  tau_obs ~ dgamma(a_obs,r_obs)
  Omega_proc ~ dwish(R,k)
  alpha ~ dbeta(1,20)
}
"
```

And the parameters estimated from the model

```{r}
## load parameters (assume known)
load("data/KFalpha.params.Rdata")
## observation error
tau_obs
## process error covariance
knitr::kable(tau_proc,col.names = states)
## process error correlation
knitr::kable(cov2cor(tau_proc),col.names = states)
## process error SD
sqrt(diag(tau_proc))
```

If you were running the KF in 'operational' mode, where new data is arriving in real time, you wouldn't run the function all at once but rather just call the KalmanAnalysis every time new data is observed, followed by KalmanForecast to make a new forecast.

```{r}
##'  Kalman Filter
##' @param  M   = model matrix
##' @param  mu0 = initial condition mean vector
##' @param  P0  = initial condition covariance matrix
##' @param  Q   = process error covariance matrix
##' @param  R   = observation error covariance matrix
##' @param  Y   = observation matrix (with missing values as NAs), time as col's
##'
##' @return list
##'  mu.f, mu.a  = state mean vector for (a)nalysis and (f)orecast steps
##'  P.f, P.a    = state covariance matrix for a and f
KalmanFilter <- function(M,mu0,P0,Q,R,Y){
  
  ## storage
  nstates = nrow(Y)  
  nt = ncol(Y)
  mu.f  = matrix(NA,nstates,nt+1)  ## forecast mean for time t
  mu.a  = matrix(NA,nstates,nt)  ## analysis mean for time t
  P.f  = array(NA,c(nstates,nstates,nt+1))  ## forecast variance for time t
  P.a  = array(NA,c(nstates,nstates,nt))  ## analysis variance for time t

  ## initialization
  mu.f[,1] = mu0
  P.f[,,1] = P0
  I = diag(1,nstates)

  ## run updates sequentially for each observation.
  for(t in 1:nt){

    ## Analysis step: combine previous forecast with observed data
    KA <- KalmanAnalysis(mu.f[,t],P.f[,,t],Y[,t],R,I)
    mu.a[,t] <- KA$mu.a
    P.a[,,t] <- KA$P.a
    
    ## Forecast step: predict to next step from current
    KF <- KalmanForecast(mu.a[,t],P.a[,,t],M,Q)
    mu.f[,t+1] <- KF$mu.f
    P.f[,,t+1] <- KF$P.f
  }
  
  return(list(mu.f=mu.f,mu.a=mu.a,P.f=P.f,P.a=P.a))
}

##' Kalman Filter: Analysis step
##' @param  mu.f = Forecast mean (vector)
##' @param  P.f  = Forecast covariance (matrix)
##' @param  Y    = observations, with missing values as NAs) (vector)
##' @param  R    = observation error covariance (matrix)
##' @param  H    = observation matrix (maps observations to states)
KalmanAnalysis <- function(mu.f,P.f,Y,R,H){
  obs = !is.na(Y) ## which Y's were observed?
  if(any(obs)){
    H <- H[obs,]                                              ## observation matrix
    K <- P.f %*% t(H) %*% solve(H%*%P.f%*%t(H) + R[obs,obs])  ## Kalman gain
    mu.a <- mu.f + K%*%(Y[obs] - H %*% mu.f)                  ## update mean
    P.a <- (1-K %*% H)*P.f                                    ## update covariance
  } else {
    ##if there's no data, the posterior is the prior
    mu.a = mu.f
    P.a = P.f
  }
  return(list(mu.a=mu.a,P.a=P.a))
}

##' Kalman Filter: Forecast Step
##' @param mu.a = analysis posterior mean (vector)
##' @param P.a  = analysis posterior covariance (matrix)
##' @param M    = model (matrix)
##' @param  Q   = process error covariance (matrix)
KalmanForecast <- function(mu.a,P.a,M,Q){
  mu.f = M%*%mu.a
  P.f  = Q + M*P.a*t(M)
  return(list(mu.f=mu.f,P.f=P.f))
}
```

Define the inputs to the function. 
KF00 indicates that this run was done with the defaults for both the process model and process error covariance.

```{r}
## log transform data
Y = log10(y)

## options for process model 
alpha = 0       ## assume no spatial flux
#alpha = 0.05    ## assume a large spatial flux
M = adj*alpha + diag(1-alpha*apply(adj,1,sum))  ## random walk with flux

## options for process error covariance
Q = tau_proc            ## full covariance matrix
#Q = diag(diag(Q))       ## diagonal covariance matrix

## observation error covariance (assumed independent)  
R = diag(tau_obs,nstates) 

## prior on first step, initialize with long-term mean and covariance
mu0 = apply(Y,1,mean,na.rm=TRUE)
P0 = cov(t(Y),use="pairwise.complete.obs")

## Run Kalman Filter
KF00 = KalmanFilter(M,mu0,P0,Q,R,Y)
```

Visualize the outputs. 
The first set of figures below shows the posterior analysis for each state through time. 
The second set shows the forecast and analysis standard deviations change through time, indicating when there is missing data in green on the bottom of the plot. 
As you can see the missing data is not synchronous across states, but the mean of the Analysis is influenced by the across-state covariances.

```{r, fig.asp=1.0}
attach(KF00)
nt = length(time)

### plot ANALYSIS mean & CI time-series
par(mfrow=c(3,1))
for(i in 1:6){
  ci = rbind(mu.a[i,]-1.96*sqrt(P.a[i,i,]),mu.a[i,]+1.96*sqrt(P.a[i,i,]))
  plot(time,mu.a[i,],ylim=range(ci,na.rm=TRUE),type='n',main=paste0(states[i],", alpha=0, original Q matrix"))
  ecoforecastR::ciEnvelope(time,ci[1,],ci[2,],col="lightBlue")
  lines(time,mu.a[i,],col=4)
  lines(time,Y[i,])
}

## plot ANALYSIS and FORECAST variance time-series
par(mfrow=c(3,1))
for(i in 1:6){
  plot(time,sqrt(P.a[i,i,]),ylim=c(0,sqrt(max(c(P.a[i,i,],P.f[i,i,])))),
       main=paste0(states[i],", alpha=0, original Q matrix"),xlab="Time",
       ylab="Std Error",type='l')
  lines(time,sqrt(P.f[i,i,1:nt]),col=2)
  points(time[is.na(Y[i,])],rep(0,nt)[is.na(Y[i,])],pch="*",col=3) ## flag's the zero's
  legend("topright",legend=c("Analysis","Forecast","NAs"),col=1:3,lty=c(1,1,NA),pch=c(NA,NA,1),cex=1.4)
}
```

Finally, to get a better idea about the dynamics of how the Kalman Filter works we can zoom in to a subset of time for one state and show the Forecast, Analysis, and observed data altogether.

```{r}
## subset time
time2 <- time[time>as.Date("2015-01-01")]
tsel <- which(time %in% time2)
n = length(time2)*2

## interleave Forecast and Analysis
mu = p = rep(NA,n)
mu[seq(1,n,by=2)] = mu.f[1,tsel]
mu[seq(2,n,by=2)] = mu.a[1,tsel]
p[seq(1,n,by=2)]  = 1.96*sqrt(P.f[1,1,tsel])
p[seq(2,n,by=2)]  = 1.96*sqrt(P.a[1,1,tsel])
ci = cbind(mu-p,mu+p)
time3 = sort(c(time2,time2+1))

## plot Forecast, Analysis, and data
plot(time3,mu,ylim=range(ci),type='l')
ecoforecastR::ciEnvelope(time3,ci[,1],ci[,2],col="lightBlue")
lines(time3,mu,lwd=2)
points(time,Y[1,])
```



### 1. KF01
* KF01: Rerun with process error set to just the diagonal matrix of Q, compare the results with the original -- what impact does including covariance in the process error have on the inference?

```{r}
## options for process error covariance
#Q = tau_proc            ## full covariance matrix
Q = diag(diag(Q))       ## diagonal covariance matrix

## Run Kalman Filter
KF01 = KalmanFilter(M,mu0,P0,Q,R,Y)
```

Visualize the outputs. 

```{r, fig.asp=1.0}
attach(KF01)
nt = length(time)

### plot ANALYSIS mean & CI time-series
par(mfrow=c(3,1))
for(i in 1:6){
  ci = rbind(mu.a[i,]-1.96*sqrt(P.a[i,i,]),mu.a[i,]+1.96*sqrt(P.a[i,i,]))
  plot(time,mu.a[i,],ylim=range(ci,na.rm=TRUE),type='n',main=paste0(states[i],", alpha=0, diagonal Q matrix"))
  ecoforecastR::ciEnvelope(time,ci[1,],ci[2,],col="lightBlue")
  lines(time,mu.a[i,],col=4)
  lines(time,Y[i,])
}

## plot ANALYSIS and FORECAST variance time-series
par(mfrow=c(3,1))
for(i in 1:6){
  plot(time,sqrt(P.a[i,i,]),ylim=c(0,sqrt(max(c(P.a[i,i,],P.f[i,i,])))),
       main=paste0(states[i],", alpha=0, diagonal Q matrix"),xlab="Time",
       ylab="Std Error",type='l')
  lines(time,sqrt(P.f[i,i,1:nt]),col=2)
  points(time[is.na(Y[i,])],rep(0,nt)[is.na(Y[i,])],pch="*",col=3) ## flag's the zero's
  legend("topright",legend=c("Analysis","Forecast","NAs"),col=1:3,lty=c(1,1,NA),pch=c(NA,NA,1),cex=1.4)
}
```

Finally, to get a better idea about the dynamics of how the Kalman Filter works we can zoom in to a subset of time for one state and show the Forecast, Analysis, and observed data altogether.

```{r}
## subset time
time2 <- time[time>as.Date("2015-01-01")]
tsel <- which(time %in% time2)
n = length(time2)*2

## interleave Forecast and Analysis
mu = p = rep(NA,n)
mu[seq(1,n,by=2)] = mu.f[1,tsel]
mu[seq(2,n,by=2)] = mu.a[1,tsel]
p[seq(1,n,by=2)]  = 1.96*sqrt(P.f[1,1,tsel])
p[seq(2,n,by=2)]  = 1.96*sqrt(P.a[1,1,tsel])
ci = cbind(mu-p,mu+p)
time3 = sort(c(time2,time2+1))

## plot Forecast, Analysis, and data
plot(time3,mu,ylim=range(ci),type='l')
ecoforecastR::ciEnvelope(time3,ci[,1],ci[,2],col="lightBlue")
lines(time3,mu,lwd=2)
points(time,Y[1,])
```

```
The mu.a (analysis mean for time t) of the KF01 is a constant when there have missing data at the beginning of the timeseries. Across all the different states, the shape of seasonal oscillation for both ANALYSIS and FORECAST variance are pretty similar.

For the sates with missing data at the beginning of timeseries, the intra-annual variability tend to be more stable when including covariance in the process error.

In the KF00 and KF01 with alpha = 0, the ANALYSIS and FORECAST variance were keep increasing when there is missing data. When data are available, ANALYSIS and FORECAST variance drop suddenly. 

```

### 2. KF11
* KF11: Rerun with alpha = 0.05 and the diagonal Q matrix. Comparing KF11 to KF01, what impact does including a spatial flux in the process model have on the inference?

```{r}
## options for process model 
#alpha = 0       ## assume no spatial flux
alpha = 0.05    ## assume a large spatial flux
M = adj*alpha + diag(1-alpha*apply(adj,1,sum))  ## random walk with flux

## Run Kalman Filter
KF11 = KalmanFilter(M,mu0,P0,Q,R,Y)
```

Visualize the outputs. 

```{r, fig.asp=1.0}
attach(KF11)
nt = length(time)

### plot ANALYSIS mean & CI time-series
par(mfrow=c(3,1))
for(i in 1:6){
  ci = rbind(mu.a[i,]-1.96*sqrt(P.a[i,i,]),mu.a[i,]+1.96*sqrt(P.a[i,i,]))
  plot(time,mu.a[i,],ylim=range(ci,na.rm=TRUE),type='n',main=paste0(states[i],", alpha=0, diagonal Q matrix"))
  ecoforecastR::ciEnvelope(time,ci[1,],ci[2,],col="lightBlue")
  lines(time,mu.a[i,],col=4)
  lines(time,Y[i,])
}

## plot ANALYSIS and FORECAST variance time-series
par(mfrow=c(3,1))
for(i in 1:6){
  plot(time,sqrt(P.a[i,i,]),ylim=c(0,sqrt(max(c(P.a[i,i,],P.f[i,i,])))),
       main=paste0(states[i],", alpha=0, diagonal Q matrix"),xlab="Time",
       ylab="Std Error",type='l')
  lines(time,sqrt(P.f[i,i,1:nt]),col=2)
  points(time[is.na(Y[i,])],rep(0,nt)[is.na(Y[i,])],pch="*",col=3) ## flag's the zero's
  legend("topright",legend=c("Analysis","Forecast","NAs"),col=1:3,lty=c(1,1,NA),pch=c(NA,NA,1),cex=1.4)
}
```

Zoom in to a subset of time for one state and show the Forecast, Analysis, and observed data altogether.

```{r}
## subset time
time2 <- time[time>as.Date("2015-01-01")]
tsel <- which(time %in% time2)
n = length(time2)*2

## interleave Forecast and Analysis
mu = p = rep(NA,n)
mu[seq(1,n,by=2)] = mu.f[1,tsel]
mu[seq(2,n,by=2)] = mu.a[1,tsel]
p[seq(1,n,by=2)]  = 1.96*sqrt(P.f[1,1,tsel])
p[seq(2,n,by=2)]  = 1.96*sqrt(P.a[1,1,tsel])
ci = cbind(mu-p,mu+p)
time3 = sort(c(time2,time2+1))

## plot Forecast, Analysis, and data
plot(time3,mu,ylim=range(ci),type='l')
ecoforecastR::ciEnvelope(time3,ci[,1],ci[,2],col="lightBlue")
lines(time3,mu,lwd=2)
points(time,Y[1,])
```

```
The mu.a (analysis mean for time t) variables of some states in the KF11 show inter-annual change when there are missing data at the beginning of the timeseries, but the seasonal oscillation (intra-annual) are relatively stable. The confidence intervals of KF11 ANALYSIS are lower than of KF01 when there have missing data, while the ANALYSIS and FORECAST variance for both KF11 and KF01 quickly drop/increase and converge to a stable value, 

After including a spatial flux in the process model, the ANALYSIS mean for the six states are more spatially correlated, but the ANALYSIS and FORECAST variance of all of states are stable. Only FORECAST variance of Maine state is increasing when there are missing data. When all the data are available, their variances drop suddenly and keep stable at a relative lower level.

```

### 3. KF10
* KF10: Rerun with alpha = 0.05 and the original Q matrix. Compare KF10 to previous runs -- what impact does including both a spatial process and a process error covariance have over their impacts individually.

```{r}
## options for process error covariance
Q = tau_proc            ## full covariance matrix

## Run Kalman Filter
KF10 = KalmanFilter(M,mu0,P0,Q,R,Y)
```

Visualize the outputs. 

```{r, fig.asp=1.0}
attach(KF10)
nt = length(time)

### plot ANALYSIS mean & CI time-series
par(mfrow=c(3,1))
for(i in 1:6){
  ci = rbind(mu.a[i,]-1.96*sqrt(P.a[i,i,]),mu.a[i,]+1.96*sqrt(P.a[i,i,]))
  plot(time,mu.a[i,],ylim=range(ci,na.rm=TRUE),type='n',main=paste0(states[i],", alpha=0, diagonal Q matrix"))
  ecoforecastR::ciEnvelope(time,ci[1,],ci[2,],col="lightBlue")
  lines(time,mu.a[i,],col=4)
  lines(time,Y[i,])
}

## plot ANALYSIS and FORECAST variance time-series
par(mfrow=c(3,1))
for(i in 1:6){
  plot(time,sqrt(P.a[i,i,]),ylim=c(0,sqrt(max(c(P.a[i,i,],P.f[i,i,])))),
       main=paste0(states[i],", alpha=0, diagonal Q matrix"),xlab="Time",
       ylab="Std Error",type='l')
  lines(time,sqrt(P.f[i,i,1:nt]),col=2)
  points(time[is.na(Y[i,])],rep(0,nt)[is.na(Y[i,])],pch="*",col=3) ## flag's the zero's
  legend("topright",legend=c("Analysis","Forecast","NAs"),col=1:3,lty=c(1,1,NA),pch=c(NA,NA,1),cex=1.4)
}
```
oom in to a subset of time for one state and show the Forecast, Analysis, and observed data altogether.

```{r}
## subset time
time2 <- time[time>as.Date("2015-01-01")]
tsel <- which(time %in% time2)
n = length(time2)*2

## interleave Forecast and Analysis
mu = p = rep(NA,n)
mu[seq(1,n,by=2)] = mu.f[1,tsel]
mu[seq(2,n,by=2)] = mu.a[1,tsel]
p[seq(1,n,by=2)]  = 1.96*sqrt(P.f[1,1,tsel])
p[seq(2,n,by=2)]  = 1.96*sqrt(P.a[1,1,tsel])
ci = cbind(mu-p,mu+p)
time3 = sort(c(time2,time2+1))

## plot Forecast, Analysis, and data
plot(time3,mu,ylim=range(ci),type='l')
ecoforecastR::ciEnvelope(time3,ci[,1],ci[,2],col="lightBlue")
lines(time3,mu,lwd=2)
points(time,Y[1,])
```

```
When there are missing data, their ANALYSIS and FORECAST variance quickly drop/increase and converge to a stable value.
Confidence intervals of KF10 are much smaller than KF00. In addition, they look a bit smaller than KF11 (at some time points).
```


* If the process model included a strong environmental covariate that correlated with the seasonal oscillation of influenza, how would this impact the Forecast and Analysis? Comment on how this would affect mu.f, mu.a, P.f, P.a, Q, and R specifically.

```
If the process model included a strong environmental covariate that correlated with the seasonal oscillation of influenza, forecast mean (mu.f) would still be relatively stable, but analysis mean (mu.a) would be more inconstantly. Forecast variance would not change too much, while analysis variance would be a bit larger. 
Process error covariance (Q) would stay barely change but observation error covariance (R) would increase.

```


* Explain conceptually what would have to change in the Kalman Filter function if we replaced the current linear model with a nonlinear process model, in order to turn it into an Extended Kalman Filter function.
```
First, the linear assumption of the forecast would be reconstructed to mu.f = f(mu.a).
Second, the variance would be updated or approximated using Taylor Series expansion. I would replace M (model matrix) to F matrix. The F matrix would be calculated by a Jacobian operation: F = Jacobian(dmu.f.i/dx.j)
```
