---
title: "Assginment_07"
author: "Yingtong Zhang"
date: "3/23/2019"
output: html_document
---

By default this code is set to run with a small number of years (15), and a much too low number of MCMC iterations (500), just so that the code with "knit" quickly initially. For your analyses you should obviously increase these -- I found that convergence was adequate with around 20,000 samples, though I probably would run 10x longer than that for a publishable analysis. However, such an analysis would take hours to run.

Assignment:

1. Run the model initially with random effects off.

```{r}
if(!require(PEcAn.data.land)){
  library(devtools)
  install.packages(c("digest","dplR","PeriodicTable"),repos = "https://cloud.r-project.org")
  devtools::install_github("PecanProject/pecan/base/logger")
  devtools::install_github("PecanProject/pecan/base/remote")
  devtools::install_github("PecanProject/pecan/base/utils")
  devtools::install_github("PecanProject/pecan/base/db")
  devtools::install_github("PecanProject/pecan/modules/data.land")
  library(PEcAn.data.land)
}
library(rjags)
library(ecoforecastR)
```


```{r}
## 1. Read tree data
trees <- read.csv("data/H2012AdultFieldData.csv")

## 2. Read tree ring data
rings <- Read_Tucson("data/TUCSON/")

## 3. merge inventory and tree ring data, extract most recent nyears
combined <- matchInventoryRings(trees,rings,nyears=15)

## take a look at the first few rows of data to see the structure
knitr::kable(combined[1:5,])

## 4. organize data into a list
data <- buildJAGSdata_InventoryRings(combined)

# make sure to take a look at all the priors!
str(data)

```


```{r, fig.asp=0.8}

n.iter = 20000

## this code fuses forest inventory data with tree growth data (tree ring or dendrometer band)
## for the same plots. Code is a rewrite of Clark et al 2007 Ecol Appl into JAGS
TreeDataFusionMV = "
model{

### Loop over all individuals
for(i in 1:ni){
  
  #### Data Model: DBH
  for(t in 1:nt){
    z[i,t] ~ dnorm(x[i,t],tau_dbh)
  }
  
  #### Data Model: growth
  for(t in 2:nt){
    inc[i,t] <- x[i,t]-x[i,t-1]
    y[i,t] ~ dnorm(inc[i,t],tau_inc)
  }
  
  #### Process Model
  for(t in 2:nt){
    Dnew[i,t] <- x[i,t-1] + mu
    x[i,t]~dnorm(Dnew[i,t],tau_add)
  }
  
  x[i,1] ~ dnorm(x_ic,tau_ic)

}  ## end loop over individuals
  
  #### Priors
  tau_dbh ~ dgamma(a_dbh,r_dbh)
  tau_inc ~ dgamma(a_inc,r_inc)
  tau_add ~ dgamma(a_add,r_add)
  mu ~ dnorm(0.5,0.5)
}"

  ## state variable initial condition (subtract observed diameter increments off from the observed diameter)
  z0 = t(apply(data$y,1,function(y){-rev(cumsum(rev(y)))})) + data$z[,ncol(data$z)] 
  
  ## JAGS initial conditions
  nchain = 3
  init <- list()
  for(i in 1:nchain){
    y.samp = sample(data$y,length(data$y),replace=TRUE)
    init[[i]] <- list(x = z0,tau_add=runif(1,1,5)/var(diff(y.samp),na.rm=TRUE),
                      tau_dbh=1,tau_inc=500,tau_ind=50,tau_yr=100,ind=rep(0,data$ni),year=rep(0,data$nt))
  }
  
  ## compile JAGS model
  j.model   <- jags.model (file = textConnection(TreeDataFusionMV),
                           data = data,
                           inits = init,
                           n.chains = 3)
  ## burn-in
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("tau_add","tau_dbh","tau_inc","mu","tau_ind","tau_yr"),
                              n.iter = min(n.iter,2000))
  plot(jags.out)
  
  ## run MCMC
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("x","tau_add","tau_dbh","tau_inc","mu",
                                                 "tau_ind","tau_yr","ind","year"),
                              n.iter = n.iter)
```


```{r, fig.height=8}
  #### Diagnostic plots
  
  ### DBH
  layout(matrix(1:8,4,2))
  out <- as.matrix(jags.out)
  x.cols = which(substr(colnames(out),1,1)=="x")   ## which columns are the state variable, x
  ci <- apply(out[,x.cols],2,quantile,c(0.025,0.5,0.975))
  ci.names = parse.MatrixNames(colnames(ci),numeric=TRUE)
  
  smp = c(sample.int(data$ni,3),49)  ## I've rigged the sampling to make sure you see tree 49!
  for(i in smp){
    sel = which(ci.names$row == i)
    plot(data$time,ci[2,sel],type='n',ylim=range(ci[,sel],na.rm=TRUE),ylab="DBH (cm)",main=i)
    ciEnvelope(data$time,ci[1,sel],ci[3,sel],col="lightBlue")
    points(data$time,data$z[i,],pch="+",cex=1.5)
  }
  
  ## growth
  for(i in smp){
    sel = which(ci.names$row == i)
    inc.mcmc = apply(out[,x.cols[sel]],1,diff)
    inc.ci = apply(inc.mcmc,1,quantile,c(0.025,0.5,0.975))*5
    
    plot(data$time[-1],inc.ci[2,],type='n',ylim=range(inc.ci,na.rm=TRUE),ylab="Ring Increment (mm)")
    ciEnvelope(data$time[-1],inc.ci[1,],inc.ci[3,],col="lightBlue")
    points(data$time,data$y[i,]*5,pch="+",cex=1.5,type='b',lty=2)
  }
```


```{r}
  ## process model
  vars = (1:ncol(out))[-c(which(substr(colnames(out),1,1)=="x"),grep("tau",colnames(out)),
                          grep("year",colnames(out)),grep("ind",colnames(out)))]
  par(mfrow=c(1,1))
  for(i in vars){
    hist(out[,i],main=colnames(out)[i])
  }
  if(length(vars)>1) pairs(out[,vars])

  ## Standard Deviations
  par(mfrow=c(2,3))
  prec = out[,grep("tau",colnames(out))]
  for(i in 1:ncol(prec)){
    hist(1/sqrt(prec[,i]),main=colnames(prec)[i])
  }
  cor(prec)
  pairs(prec)
```


2. Rerun the model with random effects on. 

```{r, fig.asp=0.8}

n.iter = 20000

## this code fuses forest inventory data with tree growth data (tree ring or dendrometer band)
## for the same plots. Code is a rewrite of Clark et al 2007 Ecol Appl into JAGS
TreeDataFusionMV = "
model{
  
### Loop over all individuals
for(i in 1:ni){
  
  #### Data Model: DBH
  for(t in 1:nt){
    z[i,t] ~ dnorm(x[i,t],tau_dbh)
  }
  
  #### Data Model: growth
  for(t in 2:nt){
    inc[i,t] <- x[i,t]-x[i,t-1]
    y[i,t] ~ dnorm(inc[i,t],tau_inc)
  }
  
  #### Process Model
  for(t in 2:nt){
    Dnew[i,t] <- x[i,t-1] + mu + ind[i] + year[t]
    x[i,t]~dnorm(Dnew[i,t],tau_add)
  }
  
  ## individual effects
  ind[i] ~ dnorm(0,tau_ind)
  
  ## initial condition
  x[i,1] ~ dnorm(x_ic,tau_ic)
  
}  ## end loop over individuals
  
  ## year effects
  for(t in 1:nt){
    year[t] ~ dnorm(0,tau_yr)
  }
  
  
  #### Priors
  tau_dbh ~ dgamma(a_dbh,r_dbh)
  tau_inc ~ dgamma(a_inc,r_inc)
  tau_add ~ dgamma(a_add,r_add)
  tau_ind ~ dgamma(1,0.1)
  tau_yr  ~ dgamma(1,0.1)
  mu ~ dnorm(0.5,0.5)

}"

  ## state variable initial condition (subtract observed diameter increments off from the observed diameter)
  z0 = t(apply(data$y,1,function(y){-rev(cumsum(rev(y)))})) + data$z[,ncol(data$z)] 
  
  ## JAGS initial conditions
  nchain = 3
  init <- list()
  for(i in 1:nchain){
    y.samp = sample(data$y,length(data$y),replace=TRUE)
    init[[i]] <- list(x = z0,tau_add=runif(1,1,5)/var(diff(y.samp),na.rm=TRUE),
                      tau_dbh=1,tau_inc=500,tau_ind=50,tau_yr=100,ind=rep(0,data$ni),year=rep(0,data$nt))
  }
  
  ## compile JAGS model
  j.model   <- jags.model (file = textConnection(TreeDataFusionMV),
                           data = data,
                           inits = init,
                           n.chains = 3)
  ## burn-in
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("tau_add","tau_dbh","tau_inc","mu","tau_ind","tau_yr"),
                              n.iter = min(n.iter,2000))
  plot(jags.out)
  
  ## run MCMC
  jags.out   <- coda.samples (model = j.model,
                              variable.names = c("x","tau_add","tau_dbh","tau_inc","mu",
                                                 "tau_ind","tau_yr","ind","year"),
                              n.iter = n.iter)
```


```{r, fig.height=8}
  #### Diagnostic plots
  
  ### DBH
  layout(matrix(1:8,4,2))
  out <- as.matrix(jags.out)
  x.cols = which(substr(colnames(out),1,1)=="x")   ## which columns are the state variable, x
  ci <- apply(out[,x.cols],2,quantile,c(0.025,0.5,0.975))
  ci.names = parse.MatrixNames(colnames(ci),numeric=TRUE)
  
  smp = c(sample.int(data$ni,3),49)  ## I've rigged the sampling to make sure you see tree 49!
  for(i in smp){
    sel = which(ci.names$row == i)
    plot(data$time,ci[2,sel],type='n',ylim=range(ci[,sel],na.rm=TRUE),ylab="DBH (cm)",main=i)
    ciEnvelope(data$time,ci[1,sel],ci[3,sel],col="lightBlue")
    points(data$time,data$z[i,],pch="+",cex=1.5)
  }
  
  ## growth
  for(i in smp){
    sel = which(ci.names$row == i)
    inc.mcmc = apply(out[,x.cols[sel]],1,diff)
    inc.ci = apply(inc.mcmc,1,quantile,c(0.025,0.5,0.975))*5
    
    plot(data$time[-1],inc.ci[2,],type='n',ylim=range(inc.ci,na.rm=TRUE),ylab="Ring Increment (mm)")
    ciEnvelope(data$time[-1],inc.ci[1,],inc.ci[3,],col="lightBlue")
    points(data$time,data$y[i,]*5,pch="+",cex=1.5,type='b',lty=2)
  }
```


```{r}
  ## process model
  vars = (1:ncol(out))[-c(which(substr(colnames(out),1,1)=="x"),grep("tau",colnames(out)),
                          grep("year",colnames(out)),grep("ind",colnames(out)))]
  par(mfrow=c(1,1))
  for(i in vars){
    hist(out[,i],main=colnames(out)[i])
  }
  if(length(vars)>1) pairs(out[,vars])

  ## Standard Deviations
  par(mfrow=c(2,3))
  prec = out[,grep("tau",colnames(out))]
  for(i in 1:ncol(prec)){
    hist(1/sqrt(prec[,i]),main=colnames(prec)[i])
  }
  cor(prec)
  pairs(prec)
```

```{r}
  par(mfrow=c(1,1))
  ### YEAR
  year.cols = grep("year",colnames(out))
  if(length(year.cols>0)){
    ci.yr <- apply(out[,year.cols],2,quantile,c(0.025,0.5,0.975))
    plot(data$time,ci.yr[2,],type='n',ylim=range(ci.yr,na.rm=TRUE),main="Year Effect",ylab="cm")
    ciEnvelope(data$time,ci.yr[1,],ci.yr[3,],col="lightBlue")
    lines(data$time,ci.yr[2,],lty=1,lwd=2)
    abline(h=0,lty=2)
  }
  
  ### INDIV
  ind.cols= which(substr(colnames(out),1,3)=="ind")
  if(length(ind.cols)>0){
    boxplot(out[,ind.cols],horizontal=TRUE,outline=FALSE,col=combined$PLOT,main="Individual Effects By Plot",xlab="cm")
    abline(v=0,lty=2)
    ## calculate plot-level means for random effects
    tapply(apply(out[,ind.cols],2,mean),combined$PLOT,mean)
    table(combined$PLOT)
    
    spp = combined$SPP
    boxplot(out[order(spp),ind.cols],horizontal=TRUE,outline=FALSE,col=spp[order(spp)],main="Individual Effects By Species",xlab="cm")
    abline(v=0,lty=2)
    spp.code = levels(spp)[table(spp)>0]
    legend("bottomright",legend=rev(spp.code),col=rev(which(table(spp)>0)),lwd=4)
    ## calculate species-level means for random effects
    tapply(apply(out[,ind.cols],2,mean),combined$SPP,mean)
  }
```

Compare this to the previous run. What is the relative partitioning of uncertainties in the different versions of the model among observation error, process error, and the different random effects? What does the size of these effects suggest about the drivers of uncertainty in tree growth?

As _mu_ is the mean of growth rate, the histograms of mu in two runs show slightly differences on standard deviation. _mu_ is the model without random effects has relatively small standard deciation. DBH observation errors (_tau_dbh_) are quite similar in the two runs. The model with random effects has smaller means of _tau_add_ and _tau_inc_ compared to the first run. The credible intervals of four sample trees are all narrower in the second model. Specifically, tree #49 is very obvious in that the median values are more centralized and close to observations. 
The larger the size, the lower the drivers of uncertainty in tree growth.



3. Based on the diagnostics, propose an additional effect (fixed or random) to add to the model. Such an effect should plausibly chip away at a sizable fraction of the unexplained variability – you wouldn’t want to propose an effect that isn’t associated with systematic variability.

I would like to add an additional random effect to the model, which is the subplots. From the data we know that there are two major plots (6 and 1) and each major plot has four (A,B,C,D) subplots, so there are 8 subplots in total.

For these 8 subplots, the tree growth could be related to the soil moisture, soil temperature, and nutirents and water supply. The differences of those drivers may bacuase that the radiation and topography (sunny/shade slope) have combined effect on the soil temperature change. The current tau_add and tau_ind might decrease when have his additional effect.



4. Explain any additional exploratory analyses you would perform (e.g. plotting your proposed covariate against one of the random effects).
I would have the plot of covariances of every two of tau_inc, tau_add, tau_ind.



5. Write the JAGS code that would fit the proposed model (note: you don’t have to track down additional covariate data or run this model, just propose the code)

```
model{
  ### loop over all subplots
  for(s in 1:ns){
  
    ### Loop over all individuals within a certain subplot
    for(i in 1:ni){
      
      #### Data Model: DBH
      for(t in 1:nt){
        z[i,t,sp] ~ dnorm(x[i,t,sp],tau_dbh)
      }
      
      #### Data Model: growth
      for(t in 2:nt){
        inc[i,t,sp] <- x[i,t,sp]-x[i,t-1,sp]
        y[i,t,sp] ~ dnorm(inc[i,t,sp],tau_inc)
      }
      
      #### Process Model
      for(t in 2:nt){
        Dnew[i,t,sp] <- x[i,t-1,sp] + mu + ind[i] + year[t] + subplot[p]
        x[i,t,p]~dnorm(Dnew[i,t,p],tau_add)
      }
      
      ## individual effects
      ind[i] ~ dnorm(0,tau_ind)
      
      ## initial condition
      x[i,1,p] ~ dnorm(x_ic,tau_ic)
    }  ## end loop over individuals within a certain subplot
    
    ## subplot effects
    subplot[i] ~ dnorm(0,tau_sp)
  } ## end loop over subplots
  
  ## year effects
  for(t in 1:nt){
    year[t] ~ dnorm(0,tau_yr)
  }
  
  #### Priors
  tau_dbh ~ dgamma(a_dbh,r_dbh)
  tau_inc ~ dgamma(a_inc,r_inc)
  tau_add ~ dgamma(a_add,r_add)
  tau_ind ~ dgamma(1,0.1)
  tau_yr  ~ dgamma(1,0.1)
  tau_sp~ dgamma(1,0.1)
  mu ~ dnorm(0.5,0.5)
  
}
```


